---
title: "Homework-6-Q1"
output: 
  pdf_document:
    latex_engine: xelatex
---

BIOS507 Spring 2025 | Dr Lukemire | Elizabeth Nemeti
Due: April 7 2025

**Problem 1.**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(patchwork)
```

A person’s muscle mass is expected to decrease with age. To explore this relationship in women, a nutritionist randomly selected 15 women from each 10-year age group, beginning with age 40 and ending with age 79. The dataset is “musclemass.txt”. Note: These questions are adapted from questions 8.4 and 8.5 in the textbook.

- **Y (response)**:  Mass
- **X1 (identifier)**: Age

```{r}
data_path = "/Users/elizabethnemeti/Documents/GitHub/BIOS507-Coursework/Homeworks/Homework-6/"
data_file <- file.path(data_path, "musclemass.txt")
MUSCLEMASS_data <- read.delim(data_file, header = TRUE)

head(MUSCLEMASS_data)
str(MUSCLEMASS_data)
```
**Part A. Fit a quadratic regression model with centered age. Plot the fitted regression function and the data. Does the quadratic regression function visually appear to be a good fit here? Report the R2 value.**

```{r}

```


**Part B. Conduct an overall test for model fit using α = 0.05. What are your findings?**

```{r}

```

**Part C. Test whether the quadratic term can be dropped from the model using α = 0.05. What are your findings?**

```{r}

```

**Part D. Find and interpret a 95% confidence interval for the mean muscle mass for a women who is age 50.**

```{r}

```

**Part E. Find and interpret a 95% prediction interval for the muscle mass for a women who is age 50.**

```{r}

```

**Part F. Fit the third-order model and test for the significance of the cubic term (using α = 0.05) **

```{r}

```

