---
title: "Homework-6-Q2"
output: 
  pdf_document:
    latex_engine: xelatex
---

BIOS507 Spring 2025 | Dr Lukemire | Elizabeth Nemeti
Due: April 7 2025

**Problem 2.**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(patchwork)
```

A researcher is interested in understanding the relationship between income (measured in thousands of dollars per year) and the amount of money spent yearly on food (also measured in thousands of dollars). The researcher has collected data on these two variables, see income.csv. Conduct an analysis to model the relationship between these two variables, using the food expenditures the outcome. Be sure to state all assumptions, check their validity, describe any statistical tests you perform, and perform any required remedial measures. 

- **Y (response)**:  Mass
- **X1 (identifier)**: Age

```{r}
data_path = "/Users/elizabethnemeti/Documents/GitHub/BIOS507-Coursework/Homeworks/Homework-6/"
data_file <- file.path(data_path, "food_income.csv")
FOODINCOME_data <- read.csv(data_file, header = TRUE)

head(FOODINCOME_data)
str(FOODINCOME_data)
```


```{r}

```

