---
title: "Homework-5-Q2"
output: 
  pdf_document:
    latex_engine: xelatex
---

BIOS507 Spring 2025 | Dr Lukemire | Elizabeth Nemeti
Due: March 24 2025

**Problem 2.**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyverse)
```

Public health researchers are investigating how lifestyle factors like sleep duration and physical activity influence cholesterol levels, and whether this relationship differs based on dietary habits. Cholesterol levels are a key indicator of cardiovascular health, with high levels increasing the risk of heart disease. 

The study categorizes participants into three dietary patterns: **Plant-Based Diets, Balanced Diets, and High-Meat Diet**. The hypothesis is that increased sleep and physical activity are generally associated with lower cholesterol levels, but the magnitude of these effects differs based on dietary habits, with plant-based eaters potentially benefiting more due to better metabolic profiles, while high-meat consumers may show a weaker response. 

The dataset is diet_sleep_exercise_cholesterol.csv. Note that this is simulated data for this example homework problem and does not represent a real study.

- **Y (response)**:   cholesterol (high levels increasing the risk of heart disease)
- **X1 (predictor)**: sleepHours (sleep duration)
- **X2 (predictor)**: activity (physical activity)
- **X3 (predictor)**: diet (three dietary patterns: High-Meat; Plant-Based; Balanced)

```{r}
data_path = "/Users/elizabethnemeti/Documents/GitHub/BIOS507-Coursework/Homeworks/Homework-5/"
data_file <- file.path(data_path, "diet_sleep_exercise_cholesterol.csv")
health_data <- read.csv(data_file, header = TRUE)

head(health_data)
str(health_data)
```
**1. Fit a model (Model 1) that includes all main effects and two-factor interactions. Conduct a test for the presence of the sleep duration × physical activity interaction. Conduct a test for the presence of the diet × physical activity interaction. **

```{r}
health_data$diet <- as.factor(health_data$diet)

health_model <- lm(cholesterol ~ sleepHours * activity * diet, data = health_data)
summary(health_model)
```

**Conduct a test for the presence of the sleep duration × physical activity interaction.**
Full model: E[cholesterol] = ß0 + ß1(sleepHours) + ß2(activity) + ß3(diet) + ß4(sleepHours)(activity) + ß5(diet)(activity) + ß6(sleepHours)(diet)

Reduced model: E[cholesterol] = ß0 + ß1(sleepHours) + ß2(activity) + ß3(diet) + ß5(diet)(activity) + ß6(sleepHours)(diet)

```{r}
full_model_a <- lm(cholesterol ~ sleepHours + activity + diet + sleepHours:activity + diet:activity + sleepHours:diet, data = health_data)
reduced_model_a <- lm(cholesterol ~ sleepHours + activity + diet + diet:activity + sleepHours:diet, data = health_data)

anova_result_a <- anova(reduced_model_a, full_model_a)
print(anova_result_a)
```
**Conduct a test for the presence of the diet × physical activity interaction.**
Full model: E[cholesterol] = ß0 + ß1(sleepHours) + ß2(activity) + ß3(diet) + ß4(sleepHours)(activity) + ß5(diet)(activity) + ß6(sleepHours)(diet)

Reduced model: E[cholesterol] = ß0 + ß1(sleepHours) + ß2(activity) + ß3(diet) + ß4(sleepHours)(activity) + ß6(sleepHours)(diet)

```{r}
full_model_b <- lm(cholesterol ~ sleepHours + activity + diet + sleepHours:activity + diet:activity + sleepHours:diet, data = health_data)
reduced_model_b <- lm(cholesterol ~ sleepHours + activity + diet + sleepHours:activity + sleepHours:diet, data = health_data)

anova_result_b <- anova(reduced_model_b, full_model_b)
print(anova_result_b)
```
**2. Fit a model (Model 2) that includes all main effects, but only the diet × physical activity interaction. Create a conditional effects plot based on this model that demonstrates the interaction between physical activity and diet on cholesterol. Be sure to provide a written description of the pattern that you observe. **

```{r}
# just the diet × physical activity interaction
model_2 <- lm(cholesterol ~ sleepHours + activity + diet + diet:activity, data = health_data)

model_2_preds <- predict(model_2, newdata = health_data) # getting cholesterol predictions

# create new df for plotting, so we can better group by diet
ggplot_df <- data.frame(
  activity = health_data$activity,
  diet = health_data$diet,
  predicted_cholesterol = model_2_preds
)

ggplot(ggplot_df, aes(x = activity, y = predicted_cholesterol, color = diet)) +
  geom_line(size = 0.5) +
  labs(x = "Physical Activity",
       y = "Predicted Cholesterol (Y)",
       color = "Diet Type") +
  theme_minimal()
```
In the above plot, we can distinguish three distinct slopes for each of the 3 diet types: balanced, high-meat, and plant-based. This pattern indicates a significant interaction between diet and physical activity. The plant-based diet (blue slope) shows the steepest decline in predicted cholesterol levels as physical activity increases, suggesting that physical activity has the strongest effect on cholesterol reduction for those following a plant-based diet. In contrast, individuals following a high-meat diet (green slope) or balanced diet (orange slope) experience a weaker effect of physical activity on cholesterol reduction, as their slopes decline less sharply. All three slopes show fluctuations, which could be expected in a system as complex as cholesterol regulation. The intercept for the high-meat diet is the highest, suggesting that at zero physical activity, individuals on a high-meat diet are predicted to have the highest cholesterol levels of the three diets. Conversely, the plant-based diet starts with the lowest predicted cholesterol at zero physical activity, suggesting it may contribute to a better metabolic profile. 
