---
title: "Homework-5-Q2"
output: 
  pdf_document:
    latex_engine: xelatex
---

BIOS507 Spring 2025 | Dr Lukemire | Elizabeth Nemeti
Due: March 24 2025

**Problem 2.**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyverse)
```

Public health researchers are investigating how lifestyle factors like sleep duration and physical activity influence cholesterol levels, and whether this relationship differs based on dietary habits. Cholesterol levels are a key indicator of cardiovascular health, with high levels increasing the risk of heart disease. 

The study categorizes participants into three dietary patterns: **Plant-Based Diets, Balanced Diets, and High-Meat Diet**. The hypothesis is that increased sleep and physical activity are generally associated with lower cholesterol levels, but the magnitude of these effects differs based on dietary habits, with plant-based eaters potentially benefiting more due to better metabolic profiles, while high-meat consumers may show a weaker response. 

The dataset is diet_sleep_exercise_cholesterol.csv. Note that this is simulated data for this example homework problem and does not represent a real study.

- **Y (response)**:  cholesterol (high levels increasing the risk of heart disease)
- **X1 (predictor)**: sleepHours (sleep duration)
- **X2 (predictor)**: activity (physical activity)
- **X3 (predictor)**: diet (three dietary patterns: High-Meat; Plant-Based; Balanced)

```{r}
data_path = "/Users/elizabethnemeti/Documents/GitHub/BIOS507-Coursework/Homeworks/Homework-5/"
data_file <- file.path(data_path, "diet_sleep_exercise_cholesterol.csv")
health_data <- read.delim(data_file, header = TRUE)

head(health_data)
str(health_data)
```
**1. a. Fit a model (Model 1) that includes all main effects and two-factor interactions. b.Conduct a test for the presence of the sleep duration × physical activity interaction. c. Conduct a test for the presence of the diet × physical activity interaction. **



**2. d. Fit a model (Model 2) that includes all main effects, but only the diet × physical activity interaction. e. Create a conditional effects plot based on this model that demonstrates the interaction between physical activity and diet on cholesterol. f. Be sure to provide a written description of the pattern that you observe. **


