---
title: "Homework-5-Q1"
output: 
  pdf_document:
    latex_engine: xelatex
---

BIOS507 Spring 2025 | Dr Lukemire | Elizabeth Nemeti
Due: March 24 2025

**Problem 1.**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyverse)
```

ABCD Corp is a mid-sized logistics company that prides itself on maintaining high employee satisfaction. Unfortunately, recent surveys indicate that employee job satisfaction levels vary widely depending on workload and job type. To better understand these dynamics the company has conducted a study to examine how weekly work hours influence job satisfaction for each type of employee in the company. The company has two broad categories of employees:

- **Office Workers**: Includes software developers, accountants, marketing specialists, and project managers. These roles are cognitively demanding, requiring problem-solving, meeting deadlines, and long hours of computer-based work.
- **Manual Laborers**: Includes warehouse workers, machine operators, and delivery personnel. These employees perform physically demanding tasks such as lifting, operating machinery, and moving goods across warehouses or delivery routes.

The company has collected data on number of hours worked, type of employee, and job satisfaction (0 to 100). The dataset is ABCD_job_satisfaction.csv. Note that this is simulated data for this example homework problem and does not represent a real study.

- **Y (response)**:  Job_Satisfaction
- **X1 (predictor)**: Employee_ID
- **X2 (predictor)**: Work_Hours
- **X3 (predictor)**: Job_Role

```{r}
data_path = "/Users/elizabethnemeti/Documents/GitHub/BIOS507-Coursework/Homeworks/Homework-5/"
data_file <- file.path(data_path, "ABCD_job_satisfaction.csv")
ABCD_data <- read.delim(data_file, header = TRUE)

head(ABCD_data)
str(ABCD_data)
```
**1. Fit a model that captures the relationship between job type and hours worked and the outcome of job satisfaction. Be sure to carry out all the usual steps (EDA, writing the model, etc). Interpret the coefficients, or, if easier, create a plot demonstrating the effects. **



**2. The company asks you if there is evidence that the job satisfaction for a office worker who works 40 hours per week is different than a manual laborer who works 35 hours per week. Assess this, being sure to provide a confidence interval for the quantity that you report. **



**3. The company asks you if there is evidence that the job satisfaction for a office worker who works 20 hours per week is different than a manual laborer who works 40 hours per week. Assess this, being sure to provide a confidence interval for the quantity that you report. **



