---
title: "Homework-2-R-Answers"
output: pdf_document
date: "2025-02-23"
---

BIOS507 Spring 2025 | Dr Lukemire | Elizabeth Nemeti
Due: February 24 2025

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

On Canvas, you have a data set called solar.txt containing data collected during a solar energy project at Georgia Tech. The data contain several columns, but for now we are going to focus on heat flux (column labeled Y ) measured in kilowatts and radial deflection of the deflected rays (column labeled X4) measured in milliradians. The researchers are interested in using the radial deflection to predict the heat flux.
```{r}
data_path = "/Users/elizabethnemeti/Documents/GitHub/BIOS507-Coursework/Homeworks/Homework-2/"
data_file <- file.path(data_path, "solar.txt")
solar_data <- read.table("solar.txt", header = TRUE, sep = "") # need to add sep"" so we get individual columns
```

predictor variable (x4) -> radical deflection (milliradians) 
response variable (y) -> heat flux (kilowatts)

1. What exploratory analyses should you do using the data? Conduct these and report your findings as well as any supporting figures.

Check everything looks okay with summary():
```{r}
summary(solar_data)
```
Check whether a linear relationship is an appropriate function via a scatter plot:
```{r}
solar_data_scatterplot <- solar_data %>%
  ggplot(aes(x = x4, y = y)) +
  theme_minimal(base_size = 10) +
  geom_point() + 
  xlab("radical deflection (milliradians)") +
  ylab("heat flux (kilowatts)")

solar_data_scatterplot
```

Does a linear relationship appear appropriate? -> Yes

2. Write out the assumed regression model for Y . What are your assumptions about the model error?

heat flux (kw) = ß0 + ß1(radial deflection in milliradians) + ε

Assumptions about ε:
  - the average error should be 0. If it wasn't, that means our model is typically overpredicting/underpredicting.
  - errors should look like a bell curve and follow a normal distribution
  - there should be homoscedasticity so the error isn't depending on X4
  - there should not be a pattern in the errors, they need a random spread
  
3. Fit the model using R. Write out the estimated model.

```{r}
solar_model <- lm(y ~ x4, data = solar_data)
summary(solar_model)
```
ß0 (intercept) = 607.103
ß0 (slope) = -21.402

Estimated model -> E[heat flux (kw)] = (607.103) + (-21.402) x (radial deflection in milliradians)

4. Fill out the ANOVA table for this analysis.
```{r}
anova(solar_model)
```
          Df  Sum Sq Mean Sq F value    Pr(>F)    
x4         1 10578.7   10579  69.609 5.935e-09
Residuals 27  4103.2     152     --        --
Total     28 14681.9      --     --        --

5. Interpret the R2 value for this model.


